---
title: ""
format: md
editor: visual
---

## *Simple figures*

*In this short tutorial, we will see how to create publication-grade figures without using a mouse: we will write the instructions for the computer to draw the figures.*

### *Objectives*

*After this mini-workshop, you should:*

-   *have a basic understanding of a workflow that can build figures computationally*

-   *be able to reproduce the examples on your own data*

-   *find help if you want to make new figures*

*Before you can use this independently in your research, you will need to:*

-   *learn how to import, tidy, and process data*

-   *learn how to troubleshoot your work*

-   *practice and keep learning*

### *Want to learn more?*

*This tutorial is a very brief introduction to some powerful tools for data science. Here are a few resources that you might find useful:*

-   [*Fundamentals of Data Visualization*](clauswilke.com/dataviz) *and [Data Visualization - A practical introduction](socviz.co/index.html), excellent books providing fundamental knowledge on data visualization. [Graphic methods for presenting facts](https://archive.org/details/graphicmethodsfo00brinrich/mode/1up?view=theater), a reference book on data visualization Willard c. Brinton.*

-   [*R for data science*](r4ds.hadley.nz)*, a book on how to import, tidy, transform, visualize, and model data using the tools that we introduce in this workshop.*

-   [*Open Science, Open Data, Open Source*](osodos.org)*, OSODOS, a book providing an overview over developments in open science.*

-   *Resources on [scientific color palettes](fabiocrameri.ch/colourmaps) and [color perception simulation](color-blindness.com/coblis-color-blindness-simulator).*

-   [*Weird but sometimes useful charts*](xeno.graphics) *to get some ideas.*

### *Before the workshop...*

*In order to get started quickly on the day we run this tutorial, please do the following steps - it should only take a few minutes. If you encounter issues, no problem, we can discuss them at the beginning of the workshop.*

*📝 Open [Krug et al., doi: 10.1016/j.cell.2020.10.036](https://doi.org/10.1016/j.cell.2020.10.036) and look at the figures. Would you be able to make such figures for one of your papers?*

*📝 Install the latest version of RStudio \[link\](posit.co/download/rstudio-desktop/). Note that you will be asked to install R first.*

*📝 Optional but recommended: create an account at [GithHub](github.com) and [OpenAI](chat.openai.com).*

*📝 Prepare two sets of data: (1) one data set with values from different categories, e.g. values in different conditions; (2) one data set with one value to plot against the other, e.g. height and weight of individuals.*

*💡 Good file organization and naming can be a life saver, especially if you need to look up or change a figure after some time and in a stressful situation: a paper revision, the end of your PhD... Make sure to organize files in a clear folder organization, and use meaningful names. More on this in [OSODOS](https://pfern.github.io/OSODOS/gitbook/WORKFLOWS).*

*📝 Make the plots the way you are used to using your favorite software and arrange them as a panel A and B for a publication.*

*💬 How much did you use your mouse? What are the pros and cons of this approach?*

### *Download to your computer*

*This tutorial is made as a [notebook](https://en.wikipedia.org/wiki/Notebook_interface), a convenient way to process and plot data, while at the same time documenting what you do in a single document. Everything is stored online in [GitHub](https://github.com/mvaudel/tutorials), where you can freely access and navigate code. The project is version-controlled using [git](https://en.wikipedia.org/wiki/Git), which means that you can track changes, have multiple versions on your computer, and revert to older versions, which is an essential component of [good practices in scientific data handling](https://pfern.github.io/OSODOS/gitbook/VERSIONING/).*

*📝 Open RStudio. Click on File -\> New Project..., Select Version Control -\> Git. As Repository URL paste `https://github.com/mvaudel/tutorials.git`. Use the Browse button to select where to download the files, and click Create Project.*

*📝 In the tab called Files open the folder simple_figures and open the file named simple_figures.qmd. You should now see the text of the tutorial.*

### *Install packages*

*Many packages are available for data science in different programming languages. In general, there is no one-size-fits-all package or pipeline, so it is very convenient to be able to pick and choose based on the requirements of a specific project.*

*📝 To install the packages needed for this tutorial on your computer, click on the arrow in the code cell below.*

```{r}

if (!"tidyverse" %in% installed.packages()) { # Skip if the package is already installed
  
  install.packages("tidyverse") # A set of package to handle and plot data
  
}
if (!"ggforce" %in% installed.packages()) {
  
  install.packages("ggforce")   # A package containing functions to create custom plots
  
}
if (!"scico" %in% installed.packages()) {
  
  install.packages("scico")   # A package containing functions to create custom plots
  
}
```

### *Load packages*

*Now that the packages are installed, you can load them.*

```{r}

library("tidyverse") # A set of package to handle and plot data
library("ggforce")   # A package containing functions to create custom plots
library("scico")     # A package with scientific color palettes

```

### *Values in categories*

*In this section, we will plot values that are in different categories.*

#### *Load data*

*The data we are going to plot are gene vs. RNA vs. Protein abundance correlation values obtained from multiple cancer cell lines by [Krug et al., doi: 10.1016/j.cell.2020.10.036](https://doi.org/10.1016/j.cell.2020.10.036).*

*Note that for the sake of time, the data has been imported from the supplementary files and put in a single table. You can see how in the script named [data_preparation.R](data_preparation.R) in this folder.*

```{r}

correlation_table = read.table(
  file = "resources/correlation.gz",
  header = T,
  sep = "\t"
)

```

*The table should now appear under the Environment tab, and if you click on it you will be able to inspect its content.*

*💬 Is it not strange to have all correlation values in a single column, instead of making one column for CNA vs. RNA, and one for RNA vs. Protein?*

#### *Plot the correlation values*

*A simple way to compare the correlation values between these two categories would be to make a box plot.*

```{r}

ggplot() +
  geom_boxplot(
    data = correlation_table,
    mapping = aes(
      x = comparison,
      y = r,
      fill = comparison
    ),
    alpha = 0.8
  )

```

*To have a finer view at the distribution of points, one often makes violin plots. Note how easy it is to switch between geometries.*

```{r}

ggplot() +
  geom_violin(
    data = correlation_table,
    mapping = aes(
      x = comparison,
      y = r,
      fill = comparison
    ),
    alpha = 0.8
  )
```

*These summary statistics [can hide important details](https://nightingaledvs.com/ive-stopped-using-box-plots-should-you/). An interesting trade-off is the [sina plot](https://clauswilke.com/dataviz/boxplots-violins.html#fig:lincoln-temp-sina), which shows the individual points with the jitter following the density.*

```{r}

ggplot() +
  geom_sina(
    data = correlation_table,
    mapping = aes(
      x = comparison,
      y = r,
      col = comparison
    ),
    alpha = 0.1
  )

```

*You can easily overlay the different geometries, e.g. to combine the sina and box plots.*

```{r}

ggplot() +
  geom_sina(
    data = correlation_table,
    mapping = aes(
      x = comparison,
      y = r,
      col = comparison
    ),
    alpha = 0.1
  ) +
  geom_boxplot(
    data = correlation_table,
    mapping = aes(
      x = comparison,
      y = r,
      fill = comparison
    ),
    alpha = 0.8,
    width = 0.3,
    outlier.colour = NA
  )
```

*Before showing such a plot to your colleagues, you might want to change a few things: the order of the categories, the labels, the colors, etc.*

```{r}

correlation_table$comparison = factor(correlation_table$comparison, levels = c("r_cna_rna", "r_cna_protein"))

levels(correlation_table$comparison) = c("RNA vs. Gene", "Protein vs. Gene")

ggplot() +
  theme_bw(
    base_size = 14
  ) +
  geom_hline(
    yintercept = 0
  ) +
  geom_sina(
    data = correlation_table,
    mapping = aes(
      x = comparison,
      y = r,
      col = comparison
    ),
    alpha = 0.1
  ) +
  geom_boxplot(
    data = correlation_table,
    mapping = aes(
      x = comparison,
      y = r,
      fill = comparison
    ),
    alpha = 0.8,
    width = 0.3,
    outlier.colour = NA
  ) +
  scale_y_continuous(
    name = "Spearman's R"
  ) +
  theme(
    axis.title.x = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    panel.border = element_blank(),
    legend.position = "none"
  )
```
